The tracing of trade: exploring the patterns of olive oil production and distribution from Roman Baetica

### 2020-11-05 Paper/jaccardbaetica.tex
submission ready

%\title{The markings of the trade: exploring the patterns of olive oil production in Roman Baetica}
\title{The tracing of trade: exploring the patterns of olive oil production and distribution from Roman \textit{Baetica}}

@@

Our case of study has been focused on amphoric production from 1st to 3rd AD. In particular, we want to detect patterns of olive oil production that link amphora workshops and amphoric stamps. Amphoric stamps are used to identify the presence of different groups that might share similar stamps.

@@

This work focuses on the debate about the dynamics of the distribution of olive oil market in Roman provinces. Amphoric production in \textit{Baetica} province will help us to identify links between production centres and consumption centres in order to detect an economic pattern in the commerce by analysing stamps amphorae finding in the provinces.

@@

The chronology of the workshops is wide spanning from the first to the third centuries AD\citep{millet_anforas_1998,rodriguez_baetican_1998,chic2005comercio}. 
This fact is shown by the archaeological evidence that displays a highly specialised production with a long activity \citep{remesal_anforas_2004}. Thus, the specialisation of this production made amphorae with apparently few changes in their forms, reaching a high degree of standardisation.

@@

Scholars agree that stamps are some type of identification mark, but there is no consensus on the meaning of these marks\citep{rodriguez_baetican_1998}.

@@

However, some hypotheses argue elsewhere that this same \textit{praefectura annonae} could also have organized the supply of additional goods such as olive oil to the Roman legions  \citep{remesal_annona_1986,
remesal_annona_1990}

@@

This intense consumption of olive oil also indicates a possible provincial structure designed to organize the supply of olive oil to military camps.

@@

It seems that some Baetican centres could have been assigned to a specific province for the olive oil supply to military garrisons \citep[125]{remesal_concierto}.

However, this hypothesis is hard to assess given the current lack of written evidence related to the meaning of the amphoric stamps.

Previous studies suggested that the supply to the German limes was mainly based on riverine transport, but recent works suggest that the maritime route through the Atlantic Ocean could have been more important than expected (both for \textit{Germania} and \textit{Britannia}).~\citep{remesal_germn_2010,rubio-campillo_ecology_2018}.

@@

The production area for Dressel 20 amphorae extended across three different \textit{conventus}: \textit{Hispalensis} (currently Seville, hereafter Hispalis), \textit{Cordubensis} (currently C\'ordoba, hereafter Corduba) and \textit{Astigi} (currently Écija, Sevilla, hereafter Astigi) \citep{rodriguez_economioleicola_1977,chicdatos2001,berni_millet_epigrafianforica_2008} .

@@

The analysis of stamps in consumption areas also used the CEIPAC database and filtered in the same way than the previous dataset. However, in this case, we selected the centres with more or equal than five stamps in order to maximize the analysis with more stamps in the dataset.

@@

The approach proposed here aimed at exploring links between production and consumption centres through the identification of common amphorae codes found in different sites. For this reason we measured the similarity between amphora workshops and/or consumption sites by quantifying a pairwise distance or dissimilarity index between places (i.e. to what extent the stamp codes found on these sites were different). The chosen dissimilarity measure was the Morisita-Horn index \citep{morisita_measuring_1959, horn_measurement_1966}. This method was applied to measure the dissimilarity between different samples of sets. Generally, it describes the dissimilarity between the system of two communities based on the idea of inverse correlation between diversity and species \citep{magurran_why_1988}. In our case, stamps from different sites will be used as a proxy to measure their dissimilarity by analyzing their code.

@@

Considering our dataset as a non-uniform sample, this method provides a useful tool to handle large samples with different sizes and diversity \citep{wolda_similarity_1981}. Morisita-Horn index gives a value between 0 (sites have exact stamp codes and frequency distribution) and 1 (complete difference between stamp codes). To apply Morisita-Horn, we first calculated the number of times that each stamp code appeared in an amphora workshop. 
This method allowed us to bear in mind the number of frequency of stamps for each workshop.

@@

The hierarchical clustering of the Morisita-Horn pairwise distance matrix can be seen as a dendrogram in Figure~\ref{dendro}. This visualization suggests that each amphora workshop often used unique stamps in their production system that have not been found on any other site. In fact, the similarity values are always low and a majority of stamp codes found on multiple sites are only present in two or three workshops.

@@

\begin{sidewaysfigure}[htp]
	\centering
\includegraphics[angle=180, width=\linewidth]{figs/dendro}
\caption{Dendrogram displaying the results of hierarchical clustering based on Morisita-Horn metrics for Dressel 20 workshops located in the \textit{Baetica} area. Site names are color-coded by conventus: \textit{Hispalis} (red), \textit{Astigi} (green) and \textit{Corduba} (blue)}
\label{dendro}
\end{sidewaysfigure}

@@

\begin{sidewaysfigure}[htp]
	\centering
\includegraphics[angle=180,width=\linewidth]{figs/dendrobrit5.pdf}
\caption{Dendrogram displaying the results of hierarchical clustering based on Morisita-Horn metrics for sites in \textit{Britannia}. Site names are color-coded by typology: military sites (red), civilian sites (green) and unspecified (grey)}
\label{britmap}
\end{sidewaysfigure}

@@

\begin{sidewaysfigure}[htp]
	\centering
\includegraphics[angle=180, width=\linewidth]{figs/dendroger5.pdf}
\caption{Dendrogram displaying the results of hierarchical clustering based on Morisita-Horn metrics for sites in \textit{Germania}. Site names are color-coded by typology: military sites (red), civilian sites (green) and unspecified (grey)}
\label{germap}
\end{sidewaysfigure}

@@

This could be explained such as batch systematic organisation between potters \citep{juanmorostesis}. This work method could have allowed potters to identify batches by stamping amphorae for posterior commercialization.

@@

High production of amphorae had to maintain a specialize system organization to control this production \citep[104]{juanmorostesis}.

@@

Finally, the results suggest that there is no clear organizational pattern linking production centres and/or consumption centres beyond the spatial distance. As a consequence, our analysis does not indicate the presence of a specific workshop from \textit{Baetica} supplying to some specific province.

@@

This work has used quantitative analysis to explore long-range trade of olive oil using the stamps found in Dressel 20 vessels. The most interesting result is linked to the meaning of these inscriptions. The use of specific codes seems to be almost exclusively decided by each workshop, so it could be used to identify amphorae found in consumption centers (i.e. an amphora with a given stamp was probably made on a specific workshop). This idea could explain why the similarity of stamps between workshops is generally very low. On the other hand, some code stamps have been identified in nearby workshops so this pattern suggests that the clustered workshops were somehow connected (e.g. they belonged to the same owner).

@@

To conclude, the method presented here provides a useful framework to improve our understanding of the organisation and processes that allowed the Roman Empire to manage a massive and highly specialized production of essential goods. This work has identified similarities between amphoric stamps and spatial distribution in the case of the amphoric production within the Roman Empire.

@@

This research was partly funded by the European Research Council Advanced Grant EPNet (340828). We are grateful to Simon Carrignon, Juan Moros, Ignacio Morer and Víctor M. Lozano for their useful suggestions.  
Data have been analysed and conducted in R program version 3.2.4, using the packages \textit{vegan} \citep{oksanen_vegan_2007}, \textit{ggplot2} \citep{ggplot2:_2016}. Datasets and source code are freely available at https://github.com/Mcotsar/Baetica\_stamps.

>>>>

%\title{The markings of the trade: exploring the patterns of olive oil production in Roman Baetica}
\title{The tracing of trade: exploring the patterns of olive oil production and distribution from Roman Baetica}

@@

Our case of study has been focused on amphoric production from 1st to 3rd centuries AD. In particular, we want to detect patterns of olive oil production that link amphora workshops and amphoric stamps. Amphoric stamps are used to identify the presence of different groups that might share similar stamps.

@@

This work focuses on the debate about the dynamics of the distribution of olive oil market in Roman provinces. Amphoric production in \textit{Baetica} province will help us identify links between production centres and consumption centres in order to detect an economic pattern in the commerce by analysing stamps amphorae finding in the provinces.

@@

The chronology of the workshops is wide spanning from the first to the third centuries AD~\citep{millet_anforas_1998,rodriguez_baetican_1998,chic2005comercio}. 
This fact is shown by the archaeological evidence that displays a highly specialised production with a long activity \citep{remesal_anforas_2004}. Thus, the specialisation of this production made amphorae with apparently few changes in their forms, reaching a high degree of standardisation.

@@

Scholars agree that stamps are some type of identification mark, but there is no consensus on the meaning of these marks~\citep{rodriguez_baetican_1998}.

@@

However, some hypotheses argue elsewhere that this same \textit{praefectura annonae} could also have organised the supply of additional goods such as olive oil to the Roman legions  \citep{remesal_annona_1986,
remesal_annona_1990}

@@

This intense consumption of olive oil also indicates a possible provincial structure designed to organise the supply of olive oil to military camps.

@@

It seems that some Baetican centres could have been assigned to a specific province for the olive oil supply to military garrisons \citep[125]{remesal_concierto}. However, this hypothesis is hard to assess given the current lack of written evidence related to the meaning of the amphoric stamps.

Previous studies suggested that the supply to the German limes was mainly based on riverine transport, but recent works suggest that the maritime route through the Atlantic Ocean could have been more important than expected (both for \textit{Germania} and \textit{Britannia})~\citep{remesal_germn_2010,rubio-campillo_ecology_2018}.

@@

The production area for Dressel 20 amphorae extended across three different \textit{conventus}: \textit{Hispalensis} (currently Seville, hereafter \textit{Hispalis}), \textit{Cordubensis} (currently C\'ordoba, hereafter \textit{Corduba}) and \textit{Astigi} (currently Écija, Sevilla, hereafter \textit{Astigi}) \citep{rodriguez_economioleicola_1977,chicdatos2001,berni_millet_epigrafianforica_2008} .

@@

The analysis of stamps in consumption areas also used the CEIPAC database and filtered in the same way than the previous dataset. However, in this case, we selected the centres with more or equal than five stamps in order to maximise the analysis with more stamps in the dataset.

@@

The approach proposed here aimed at exploring links between production and consumption centres through the identification of common amphorae codes found in different sites. For this reason we measured the similarity between amphora workshops and/or consumption sites by quantifying a pairwise distance or dissimilarity index between places (i.e. to what extent the stamp codes found on these sites were different). The chosen dissimilarity measure was the Morisita-Horn index \citep{morisita_measuring_1959, horn_measurement_1966}. This method was applied to measure the dissimilarity between different samples of sets. Generally, it describes the dissimilarity between the system of two communities based on the idea of inverse correlation between diversity and species \citep{magurran_why_1988}. In our case, stamps from different sites will be used as a proxy to measure their dissimilarity by analysing their code.

@@

Considering our dataset as a non-uniform sample, this method provides a useful tool to handle large samples with different sizes and diversity \citep{wolda_similarity_1981}. Morisita-Horn index gives a value between 0 (sites have exact stamp codes and frequency distribution) and 1 (complete difference between stamp codes). To apply Morisita-Horn index, we first calculated the number of times that each stamp code appeared in an amphora workshop. 
This method allowed us to bear in mind the number of frequency of stamps for each workshop.

@@

The hierarchical clustering of the Morisita-Horn pairwise distance matrix can be seen as a dendrogram in Figure~\ref{dendro}. This visualisation suggests that each amphora workshop often used unique stamps in their production system that have not been found on any other site. In fact, the similarity values are always low and a majority of stamp codes found on multiple sites are only present in two or three workshops.

@@

\begin{sidewaysfigure}[htp]
	\centering
\includegraphics[angle=180, width=\linewidth]{figs/dendro}
\caption{Dendrogram displaying the results of hierarchical clustering based on Morisita-Horn metrics for Dressel 20 workshops located in the \textit{Baetica} area. Site names are colour-coded by \textit{conventus}: \textit{Hispalis} (red), \textit{Astigi} (green) and \textit{Corduba} (blue)}
\label{dendro}
\end{sidewaysfigure}

@@

\begin{sidewaysfigure}[htp]
	\centering
\includegraphics[angle=180,width=\linewidth]{figs/dendrobrit5.pdf}
\caption{Dendrogram displaying the results of hierarchical clustering based on Morisita-Horn metrics for sites in \textit{Britannia}. Site names are colour-coded by typology: military sites (red), civilian sites (green) and unspecified (grey)}
\label{britmap}
\end{sidewaysfigure}

@@

\begin{sidewaysfigure}[htp]
	\centering
\includegraphics[angle=180, width=\linewidth]{figs/dendroger5.pdf}
\caption{Dendrogram displaying the results of hierarchical clustering based on Morisita-Horn metrics for sites in \textit{Germania}. Site names are colour-coded by typology: military sites (red), civilian sites (green) and unspecified (grey)}
\label{germap}
\end{sidewaysfigure}

@@

This could be explained such as batch systematic organisation between potters \citep{juanmorostesis}. This work method could have allowed potters to identify batches by stamping amphorae for posterior commercialisation. High production of amphorae had to maintain a specialise system organisation to control this production \citep[104]{juanmorostesis}.

@@

@@

Finally, the results suggest that there is no clear organisational pattern linking production centres and/or consumption centres beyond the spatial distance. As a consequence, our analysis does not indicate the presence of a specific workshop from \textit{Baetica} supplying to some specific province.

@@

This work has used quantitative analysis to explore long-range trade of olive oil using the stamps found in Dressel 20 vessels. The most interesting result is linked to the meaning of these inscriptions. The use of specific codes seems to be almost exclusively decided by each workshop, so it could be used to identify amphorae found in consumption centres (i.e. an amphora with a given stamp was probably made on a specific workshop). This idea could explain why the similarity of stamps between workshops is generally very low. On the other hand, some code stamps have been identified in nearby workshops so this pattern suggests that the clustered workshops were somehow connected (e.g. they belonged to the same owner).

@@

To conclude, the method presented here provides a useful framework to improve our understanding of the organisation and processes that allowed the Roman Empire to manage a massive and highly specialised production of essential goods. This work has identified similarities between amphoric stamps and spatial distribution in the case of the amphoric production within the Roman Empire.

@@

This research was partly funded by the European Research Council Advanced Grant EPNet (340828). We are grateful to Simon Carrignon, Juan Moros, Ignacio Morer and Víctor M. Lozano for their useful suggestions.  
Data have been analysed and conducted in R program version 3.2.4, using the packages \textit{vegan} \citep{oksanen_vegan_2007}, \textit{ggplot2} \citep{ggplot2:_2016}. Maps were performed using QGIS software v.3.14 `Pi'. Datasets and source code are freely available under Open licenses from https://github.com/Mcotsar/Baetica\_stamps.
